---
title: "Week 8: Multivariate Analysis II - Supervised Learning"
author: "RM"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mva.plots)
library(ropls)
```

## Load the <data:Paracetamol_urine.Rdata>

In this data you will find 1H NMR urine spectra from 5 individuals who
were given a single dose of paracetamol. The spectra were acquired at 4
time points (0, 2, 4, and 6 hours) after drug administration.

The spec_bcor spectra were preprocessed using baseline correction only
with no normalisation. The spec_pqn spectra were preprocessed using
probabilistic quotient normalization (PQN) after baseline correction.

```{r}
# Set your working directory and load the Paracetamol urine dataset
load("Paracetamol_urine.Rdata")
```

First lets look at the spectra using matspec() from mva.plots. This will
give us an idea of the data we are working with and the region of
interest (roi).

Interest: Acetaminophen glucuronide is a metabolite of paracetamol that
can be detected in urine. It typically resonates in the region of 7.0 to
7.5 ppm in the NMR spectrum.

```{r}
matspec(spec_bcor, ppm, roi = c(7.0, 7.5))

matspec(spec_pqn, ppm, roi = c(7.0, 7.5))

```

## PLS regression using ropls

```{r}
# Subset the data to the region of interest (7.0 to 7.5 ppm)
df <- spec_pqn[, which(ppm > 7.0 & ppm < 7.5)]
Y_cont <- annotation$Paracetamol.glucuronide  # Continuous response for PLS

```

```{r}
# Run PLS regression using ropls
pls_model <- opls(df, Y_cont, predI = 2, orthoI = 0)

```

### Explanation of the figure:

**1. Top-left: R²Y and Q² per component (p1, p2)**

**Axes**

X-axis: Components (latent variables) p1 = first latent variable p2 =
second latent variable

Y-axis: Model performance (0 → 1 scale)

**Bars**

R²Y (grey): Fraction of variance in Y explained by the model Measures
goodness of fit

Q²Y (black): Fraction of variance in Y predicted by cross-validation
Measures predictive ability Interpretation

High R²Y (\~0.8–0.9) → model fits well

Moderate Q² (\~0.6–0.7) → decent predictive performance

**Gap between R² and Q²:**

Small gap → good model

Large gap → possible overfitting

**2. Top-right: Permutation test (model validation)**

**Axes**

X-axis: Similarity between original Y and permuted Y

(Similarity(y, y_perm))

-   1 = real model

-   0 or negative = random permutations

Y-axis: Model performance

-   Values of R²Y (grey) and Q²Y (black)

**Points**

Each point = a permuted model

Rightmost point = original (true) model

**Parameters**

-   pR2Y = 0.05

-   pQ2 = 0.05

These are p-values

Measure whether your model is better than random

**Interpretation**

Real model should be:

-   Far right

    -   Higher Q² than permuted models

    -   Low p-values (\<0.05) → model is statistically valid

-   If permuted Q² overlaps real → overfitting risk

**3. Bottom-left: Observation diagnostics (outlier detection)**

**Axes**

X-axis: Score Distance (SD)

-   Distance of sample in score space (model space)

-   Measures how far a sample is from the center

Y-axis: Orthogonal Distance (OD)

-   Distance from model plane

-   Measures how poorly the sample is described by the model

**Threshold lines**

Horizontal dashed line → OD limit

Vertical (implicit) threshold → SD limit

**Interpretation**

-   Low SD + Low OD: good samples

-   High SD: extreme but modeled well

-   High OD: poorly modeled (potential outlier)

-   High SD + High OD: strong outliers

Subject_4_4hours → high SD → influential sample

**4. Bottom-right: Scores plot (PLS)**

**Axes**

X-axis: t1 (17%)

-   First latent variable

-   Explains 17% of X variance

Y-axis: t2 (11%)

-   Second latent variable

-   Explains 11% of X variance

**Points**

-   Each point = sample

-   Labels = sample IDs / groups

**Ellipse**

-   Confidence ellipse (usually 95%)

    -   Defines normal variation region

**Parameters** (bottom)

R2X = 0.287

-   Variance in X explained (28.7%)

R2Y = 0.955

-   Variance in Y explained (95.5%)

Q2 = 0.638

-   Predictive ability

RMSEE = 0.642

-   Root Mean Square Error of Estimation

pre = 2

-   Number of predictive components

**Interpretation**

-   Separation along t1/t2 → group differences

-   Clustering → similar metabolic profiles

-   Points outside ellipse → potential outliers

### Question: Is this model overfitting?

**1. R² vs Q² (Top-left) → Primary overfitting check**

-   **R²Y ≈ 0.95** (very high)

-   **Q² ≈ 0.64** (moderate–good)

-   **Gap ≈ 0.31**

-   Rule of thumb:

    -   Good: Δ(R² − Q²) \< 0.2–0.3

    -   Borderline: \~0.3

    -   Overfit: \> 0.3–0.4

**2. Permutation test (Top-right) → Most important validation**

-   **pR2Y = 0.05**

-   **pQ2 = 0.05**

    -   Model is **just statistically significant**

    -   Ideally you want: pQ² \< 0.01 (strong) Clear separation from
        permuted models

--\> Model is better than random but not strongly robust \### Check the
model summary

```{r}
# Check the model summary
summary(pls_model)

```

### Get predictions from the model

```{r}
# Get predictions from the model
predictions <- predict(pls_model)
```

### Plot observed vs predicted values

```{r}
# Plot observed vs predicted values
plot(Y_cont, predictions, main = "PLS: Observed vs Predicted")
abline(0, 1, col = "red")
```

OR use fitted () function to get predicted values directly from the
model

```{r}
pred <- fitted(pls_model)

plot(Y_cont, pred,
     xlab = "Actual Y",
     ylab = "Predicted Y")
abline(0,1)
```

# OPLS - DA example

```{r}
rm(list=ls())
load("~/git/reika2/BIO513-Introduction-to-Bioinformatics-and-Data-Science-2026-S1/W7/lipo_data.RData")

## remove LTRs
lipo<-lipo[which(lipo_an$sampleType=="sample"),]
lipo_an<-lipo_an[which(lipo_an$sampleType=="sample"),]

lipo_red<-lipo[which(lipo_an$sampleTimePoint %in% c("T1","T4")),]
lipo_an_red<-lipo_an[which(lipo_an$sampleTimePoint %in% c("T1","T4")),]

Y_cat <- factor(as.character(lipo_an_red$sampleTimePoint))  # Categorical response for PLS-DA

model_oplsda <- opls(lipo_red, Y_cat, predI = 2, orthoI = 0)

```

## **Model Summary**

```{r}
model_oplsda@modelDF
```

You’ll see:

### **R2X**

-   \% of variation in X explained by model --\> “How well we summarize
    X”

### **R2Y**

-   \% of variation in Y explained --\> “How well we explain class
    separation”

### **Q2**

-   Predictive ability (via cross-validation) --\> **MOST IMPORTANT for
    model validity**

------------------------------------------------------------------------

### How to explain:

-   **R2X**: how much of X we compress into components
-   **R2Y**: how well classes are separated
-   **Q2**: how well model predicts unseen data

--\> Rule of thumb:

-   Good model: high R2Y + high Q2
-   Overfit model: high R2Y but low Q2 ❗

\##**Score Plot (Samples)**

```{r}
plot(model_oplsda, typeVc = "x-score")
```

--\> Each point = sample --\> Axes = latent variables (t1, t2)

### Interpretation:

-   Separation between Control vs Disease → good model
-   Overlap → weak discrimination

## **Loading plot (Variables)**

```{r}
plot(model_oplsda, typeVc = "x-loading")
```

--\> Shows which variables contribute to components

## **VIP Scores (Variable Importance in Projection)**

```{r}
vip <- getVipVn(model_oplsda)
head(sort(vip, decreasing = TRUE))
```

### Interpretation:

-   VIP \> 1 → important variables
-   Should highlight Var1–Var10 (true signals)

“VIP scores rank variables based on their contribution to class
separation. Values above 1 are typically considered important.”

## **Regression Coefficients**

```{r}
coef <- coef(model_oplsda)
```

### Interpretation:

-   Positive → associated with Disease
-   Negative → associated with Control

------------------------------------------------------------------------

### Explain:

“These coefficients quantify how each variable contributes to predicting
class membership — this is the regression view of the model.”

------------------------------------------------------------------------

## Prediction

```{r}
pred <- fitted(model_oplsda)

```

## \*\* Confusion Matrix\*\*

```{r}
table(pred, Y_cat)
```
