---
title: 'Week 3: Intoduction to Statistics I - Univariate Analysis'
author: "Reika Masuda"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(stats) ## you need this for qqnorm()
# -------------------------------
# Base R setup
# -------------------------------
options(digits=4)

# Manual functions
skewness <- function(x) {
  n <- length(x)
  m3 <- sum((x - mean(x))^3)/n
  s3 <- sd(x)^3
  m3 / s3
}

kurtosis <- function(x) {
  n <- length(x)
  m4 <- sum((x - mean(x))^4)/n
  s4 <- sd(x)^4
  m4 / s4 - 3
}

cv <- function(x) sd(x)/mean(x)

```

# Univariate Analysis

Univariate analysis examines one variable at a time, summarizing its central tendency, dispersion, and distribution.

## Descriptive statistics

```{r}
# Dataset A – Symmetric, low variability
data_A <- c(10, 11, 9, 10, 10, 11, 9, 10, 10, 10)

# Dataset B – Right-skewed with outlier
data_B <- c(8, 9, 10, 9, 11, 10, 9, 8, 10, 40)

# Dataset C – Symmetric, high variability
data_C <- c(2, 5, 8, 11, 14, 17, 20, 23, 26, 29)

# Let's have a look at each dataset using histogram and density plot
par(mfrow = c(1, 3))  # 3 plots side-by-side

# Dataset A
hist(data_A, probability = TRUE, main = "Dataset A",
     xlab = "Values", col = "lightblue", border = "black")
lines(density(data_A), lwd = 2)
abline(v = mean(data_A), col = "red", lwd = 2, lty = 2)    # Mean
legend("topright", legend = c("Mean"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)

# Dataset B
hist(data_B, probability = TRUE, main = "Dataset B",
     xlab = "Values", col = "lightgreen", border = "black")
lines(density(data_B), lwd = 2)
abline(v = mean(data_B), col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Mean"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)

# Dataset C
hist(data_C, probability = TRUE, main = "Dataset C",
     xlab = "Values", col = "lightpink", border = "black")
lines(density(data_C), lwd = 2)
abline(v = mean(data_C), col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Mean"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)
```

### Measures of central tendency

```{r}
mean(data_A)
median(data_A)

mean(data_B)
median(data_B)

mean(data_C)
median(data_C)


par(mfrow = c(1, 3))  # 3 plots side by side

# Dataset A
hist(data_A, probability = TRUE, main = "Dataset A",
     xlab = "Values", col = "lightblue", border = "black")
lines(density(data_A), lwd = 2)
abline(v = mean(data_A), col = "red", lwd = 2, lty = 2)    # Mean
abline(v = median(data_A), col = "darkgreen", lwd = 2, lty = 3)  # Median
legend("topright", legend = c("Mean", "Median"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)

# Dataset B
hist(data_B, probability = TRUE, main = "Dataset B",
     xlab = "Values", col = "lightgreen", border = "black")
lines(density(data_B), lwd = 2)
abline(v = mean(data_B), col = "red", lwd = 2, lty = 2)
abline(v = median(data_B), col = "darkgreen", lwd = 2, lty = 3)
legend("topright", legend = c("Mean", "Median"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)

# Dataset C
hist(data_C, probability = TRUE, main = "Dataset C",
     xlab = "Values", col = "lightpink", border = "black")
lines(density(data_C), lwd = 2)
abline(v = mean(data_C), col = "red", lwd = 2, lty = 2)
abline(v = median(data_C), col = "darkgreen", lwd = 2, lty = 3)
legend("topright", legend = c("Mean", "Median"),
       col = c("red", "darkgreen"), lty = c(2, 3), lwd = 2)

```

### Measures of dispersion

```{r}


# Datasets
data_A <- c(10, 11, 9, 10, 10, 11, 9, 10, 10, 10)
data_B <- c(8, 9, 10, 9, 11, 10, 9, 8, 10, 40)
data_C <- c(2, 5, 8, 11, 14, 17, 20, 23, 26, 29)

datasets <- list("Dataset A" = data_A,
                 "Dataset B" = data_B,
                 "Dataset C" = data_C)

par(mfrow = c(1,3), mar = c(4,4,3,1), oma = c(0,0,4,0))

for(name in names(datasets)) {
  
  data <- datasets[[name]]
  
  mean_val <- mean(data)
  sd_val   <- sd(data)
  q1       <- quantile(data, 0.25)
  q3       <- quantile(data, 0.75)
  
  hist(data, probability = TRUE,
       main = name,
       xlab = "Values",
       col = "lightblue",
       border = "black")
  
  lines(density(data), lwd = 2)
  
  abline(v = mean_val, col = "red", lwd = 2, lty = 2)
  abline(v = mean_val + sd_val, col = "darkgreen", lwd = 2, lty = 3)
  abline(v = mean_val - sd_val, col = "darkgreen", lwd = 2, lty = 3)
  abline(v = q1, col = "purple", lwd = 2, lty = 4)
  abline(v = q3, col = "purple", lwd = 2, lty = 4)
}

par(xpd = NA)
legend("bottom",
       inset = -0.25,
       horiz = TRUE,
       legend = c("Mean",
                  "Mean ± SD",
                  "Q1 & Q3 (IQR)",
                  "Density"),
       col = c("red", "darkgreen", "purple", "black"),
       lty = c(2, 3, 4, 1),
       lwd = 2,
       bty = "n")

par(mfrow = c(1,1))

summary_table <- data.frame(
  Dataset = names(datasets),
  Mean = sapply(datasets, mean),
  Median = sapply(datasets, median),
  Variance = sapply(datasets, var),
  SD = sapply(datasets, sd),
  IQR = sapply(datasets, IQR),
  CV = sapply(datasets, function(x) sd(x)/mean(x))
)

round(summary_table[,-1], 2)

```

## Probability distributions

```{r}
# Datasets
data_A <- c(10, 11, 9, 10, 10, 11, 9, 10, 10, 10)
data_B <- c(8, 9, 10, 9, 11, 10, 9, 8, 10, 40)
data_C <- c(2, 5, 8, 11, 14, 17, 20, 23, 26, 29)

datasets <- list("Dataset A" = data_A,
                 "Dataset B" = data_B,
                 "Dataset C" = data_C)
colors <- c("Dataset A"="blue", "Dataset B"="red", "Dataset C"="green")

# Set up layout: 1 row, 3 columns, space for legend
par(mfrow=c(1,3), mar=c(4,2,3,1), oma=c(0,0,4,0))

for(name in names(datasets)){
  
  data <- datasets[[name]]
  mean_val <- mean(data)
  sd_val   <- sd(data)
  q1 <- quantile(data, 0.25)
  q3 <- quantile(data, 0.75)
  cv_val <- sd_val / mean_val
  
  # Dot plot (vertical jitter)
  y_jitter <- runif(length(data), -0.05, 0.05)
  plot(data, y_jitter, pch=19, col=colors[name], 
       xlim=c(0, max(unlist(datasets))+5), ylim=c(-0.2,0.2),
       ylab="", xlab="Value", main=name, yaxt='n', cex=1.5)
  
  # Mean line
  abline(v=mean_val, col="black", lty=2, lwd=2)
  
  # SD lines (mean ± SD)
  abline(v=mean_val + sd_val, col="darkgreen", lty=3, lwd=2)
  abline(v=mean_val - sd_val, col="darkgreen", lty=3, lwd=2)
  
  # IQR lines
  abline(v=q1, col="purple", lty=4, lwd=2)
  abline(v=q3, col="purple", lty=4, lwd=2)
  
  # CV annotation
  text(mean_val, 0.15, paste0("CV=", round(cv_val,2)), cex=0.9)
}

# Super title
mtext("Spread Measures: SD, IQR & CV Across Datasets", outer=TRUE, cex=1.5)

# Shared legend below plots
par(xpd=NA)
legend("bottom", inset=-0.25, horiz=TRUE,
       legend=c("Mean", "Mean ± SD", "Q1 & Q3 (IQR)", "Dots"),
       col=c("black","darkgreen","purple","black"),
       lty=c(2,3,4,NA),
       pch=c(NA, NA, NA, 19),
       lwd=2,
       bty="n")

# Reset layout
par(mfrow=c(1,1))
```

```{r}
# Sample data
set.seed(123)
x <- c(rnorm(50, 0, 1), 3.5, -3)  # normal data + outliers

# Basic boxplot without axes
boxplot(x, horizontal = TRUE, col = "salmon", border = "purple",
        main = "Annotated Boxplot: IQR and Outliers", axes = FALSE)

# Add axis
axis(1)

# Compute quartiles
Q1 <- quantile(x, 0.25)
Q3 <- quantile(x, 0.75)
Median <- median(x)
IQR_val <- IQR(x)

# Add IQR line
arrows(Q1, 1.2, Q3, 1.2, code = 3, angle = 90, length = 0.1)
text((Q1 + Q3)/2, 1.3, "Interquartile Range (IQR)")

# Add median line label
abline(v = Median, col = "yellow", lwd = 2)
text(Median, 1.0, "Median", col = "yellow")

# Add Q1 and Q3 labels
text(Q1, 0.8, "Q1\n(25th percentile)", pos = 2)
text(Q3, 0.8, "Q3\n(75th percentile)", pos = 4)

# Add outlier labels
outliers <- boxplot.stats(x)$out
points(outliers, rep(1, length(outliers)), col = "green", pch = 16)
text(outliers[1], 1.1, "Outlier", col = "green")
text(outliers[2], 1.1, "Outlier", col = "green")

# Add whisker min/max annotations
min_whisker <- min(x[x >= (Q1 - 1.5*IQR_val)])
max_whisker <- max(x[x <= (Q3 + 1.5*IQR_val)])
text(min_whisker, 0.6, paste0('"Minimum"\n(Q1 - 1.5*IQR)'), pos = 2)
text(max_whisker, 0.6, paste0('"Maximum"\n(Q3 + 1.5*IQR)'), pos = 4)
```

## Shape of distribution

### Skewness

```{r}
# Load required library
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate sample data for different skewness
negative_skew <- rbeta(1000, 5, 1)    # Left-skewed
zero_skew     <- rnorm(1000, mean=0)  # Symmetric (normal)
positive_skew <- rbeta(1000, 1, 5)    # Right-skewed

# Combine data into a single data frame
data <- data.frame(
  value = c(negative_skew, zero_skew, positive_skew),
  skew_type = factor(rep(c("Negative Skew", "Zero Skew", "Positive Skew"), each=1000),
                     levels=c("Negative Skew", "Zero Skew", "Positive Skew"))
)

# Calculate skewness for each group
skew_labels <- aggregate(value ~ skew_type, data, function(x) round(skewness(x), 2))
# Update facet labels to include skewness
levels(data$skew_type) <- paste0(levels(data$skew_type), 
                                 " (Skewness = ", skew_labels$value, ")")

# Create density plot
ggplot(data, aes(x=value, fill=skew_type)) +
  geom_density(alpha=0.5) +
  facet_wrap(~skew_type, ncol=1) +
  theme_minimal() +
  labs(title="Illustration of Skewness",
       x="Data Values",
       y="Density") +
  scale_fill_manual(values=c("steelblue", "darkgreen", "tomato")) +
  theme(legend.position="none",
        strip.text = element_text(size=12, face="bold"))


```

### Kurtosis

```{r}
set.seed(123)

# Generate sample data
low_kurtosis  <- runif(1000, min=-3, max=3)    # Flat, light tails
normal_data   <- rnorm(1000, mean=0, sd=1)    # Normal, medium kurtosis
high_kurtosis <- rt(1000, df=3)               # Heavy tails, high kurtosis

# Combine into data frame
data <- data.frame(
  value = c(low_kurtosis, normal_data, high_kurtosis),
  kurtosis_type = factor(rep(c("Low Kurtosis", "Normal", "High Kurtosis"), each=1000),
                         levels=c("Low Kurtosis", "Normal", "High Kurtosis"))
)

# Compute kurtosis for labels
kurt_labels <- aggregate(value ~ kurtosis_type, data, function(x) round(kurtosis(x), 2))
kurt_labels$label <- paste0(kurt_labels$kurtosis_type, " (Kurtosis = ", kurt_labels$value, ")")

# Merge labels back to data for faceting
data <- merge(data, kurt_labels[, c("kurtosis_type", "label")], by="kurtosis_type")

# Plot
ggplot(data, aes(x=value, fill=kurtosis_type)) +
  geom_density(alpha=0.5) +
  facet_wrap(~label, ncol=1) +
  labs(title="Illustration of Kurtosis",
       x="Data Values",
       y="Density") +
  scale_fill_manual(values=c("skyblue", "darkgreen", "tomato")) +
  theme_minimal() +
  theme(legend.position="none",
        strip.text = element_text(size=12, face="bold"))
```

#### quick Environment cleaning

```{r}
rm(list=ls())
```

## Assessing Normality:

```{r}
# -------------------------------
# Example R Script: Assessing Normality
# -------------------------------
# Sample data: slightly skewed
set.seed(123)
data <- c(45, 50, 52, 47, 55, 60, 49, 48, 53, 51, 46, 70, 44, 65, 50, 48, 52, 49, 55, 47)
```

### histogram

```{r}
# -------------------------------
# 1. Histogram: Shows the frequency distribution of your data. Good for visualizing skewness.
# -------------------------------
hist(data, 
     breaks = 10, 
     col = "skyblue", 
     main = "Histogram of Sample Data",
     xlab = "Values",
     ylab = "Frequency")
```

### density plot

```{r}
# -------------------------------
# 2. Density Plot: Smoothed version of the histogram. Highlights the shape of the distribution.
# -------------------------------
plot(density(data), 
     main = "Density Plot of Sample Data", 
     xlab = "Values",
     ylab = "Density", 
     col = "blue", 
     lwd = 2)
polygon(density(data), col = rgb(0.2,0.5,0.8,0.3)) # shaded area
```

### Q-Q Plot

```{r}
# -------------------------------
# 3. Q-Q Plot: Compares your sample quantiles to a theoretical normal distribution. Deviations from the diagonal indicate non-normality.
# -------------------------------
qqnorm(data, main = "Q-Q Plot of Sample Data")
qqline(data, col = "red", lwd = 2)
```

### Shapiro-Wilk Test

```{r}
# -------------------------------
# 4. Shapiro–Wilk Test: Statistical test for normality. p < 0.05 → data is not normal.
# -------------------------------
shapiro_result <- shapiro.test(data)
shapiro_result

# Print result nicely
cat("Shapiro-Wilk Test:\n")
cat("W =", round(shapiro_result$statistic,3), ", p-value =", round(shapiro_result$p.value,4), "\n")
```

### Kolmogorov-Smirnov test (NOT DONE YET)

```{r}
# -------------------------------
# 5. Kolmogorov-Smirnov test:
# -------------------------------

set.seed(101)
# Example 1: Normal data
x_normal <- rnorm(50, mean=100, sd=15)

# K–S test against normal distribution
# We need to standardize data first: ks.test compares CDFs
ks_normal <- ks.test(scale(x_normal), "pnorm")
ks_normal

# Interpretation:
# - Null hypothesis (H0): sample comes from the reference distribution (here, normal)
# - p-value > 0.05 → do not reject H0 (data may be normal)
# - p-value < 0.05 → reject H0 (data differs significantly from normal)

# Visual check: histogram and Q-Q plot
par(mfrow=c(1,2))
hist(x_normal, breaks=10, probability=TRUE, col="lightblue", main="Histogram")
curve(dnorm(x, mean=mean(x_normal), sd=sd(x_normal)), col="red", add=TRUE, lwd=2)
qqnorm(x_normal); qqline(x_normal, col="red")
par(mfrow=c(1,1))

# -----------------------------------------
# Example 2: Skewed data
x_skewed <- c(rpois(20, 2), rpois(20, 10))

ks_skew <- ks.test(scale(x_skewed), "pnorm")
ks_skew

# Visual check
par(mfrow=c(1,2))
hist(x_skewed, breaks=10, col="lightgreen", main="Histogram - Skewed")
qqnorm(x_skewed); qqline(x_skewed, col="red")
par(mfrow=c(1,1))
```

```{r}
# Example data
set.seed(123)
data <- rnorm(20, mean = 50, sd = 10)

# Q-Q plot
qqnorm(data); qqline(data)

# Shapiro-Wilk test
shapiro.test(data)


# Example of non-normal data (right-skewed)
set.seed(123)
data_skew <- rexp(20, rate = 0.1)  # Exponential distribution

# Q-Q plot
qqnorm(data_skew); qqline(data_skew)

# Shapiro-Wilk test
shapiro.test(data_skew)

```

#### quick Environment cleaning

```{r}
rm(list=ls())

```

## Hypothesis testing (2 groups): parametric vs non-parametric

```{r}
# -------------------------------
# R Script: Two-Group Hypothesis Testing with Plots
# -------------------------------

# Load libraries
library(ggplot2)
library(dplyr)
```

### Parametric

#### Independent t-test

```{r}
# -------------------------------
# 1. Independent t-test (Parametric)
# -------------------------------
set.seed(123)
group_A <- c(45, 50, 52, 47, 55)
group_B <- c(60, 62, 58, 65, 63)

# Combine data for plotting
independent_df <- data.frame(
  Value = c(group_A, group_B),
  Group = factor(rep(c("A", "B"), each = 5))
)

# Boxplot
ggplot(independent_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Independent t-test: Group A vs B", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

# Independent t-test
ind_ttest <- t.test(group_A, group_B, var.equal = TRUE)
ind_ttest
```

#### Paired t-test

```{r}
# -------------------------------
# 2. Paired t-test (Parametric)
# -------------------------------
before <- c(120, 125, 130, 118, 122)
after  <- c(115, 120, 128, 110, 118)

paired_df <- data.frame(
  Subject = factor(1:5),
  Before = before,
  After = after
)

# Paired plot (lines connecting before/after)
paired_df_long <- paired_df %>%
  tidyr::pivot_longer(cols = c(Before, After), names_to = "Time", values_to = "Value")

ggplot(paired_df_long, aes(x = Time, y = Value, group = Subject)) +
  geom_line(color = "gray") +
  geom_point(aes(color = Time), size = 3) +
  labs(title = "Paired t-test: Before vs After", y = "Value") +
  theme_minimal()

# Paired t-test
paired_ttest <- t.test(before, after, paired = TRUE)
paired_ttest

# -------------------------------
# 3. Mann–Whitney U test (Non-parametric, independent)
# -------------------------------
# Same as group_A and group_B
mann_whitney <- wilcox.test(group_A, group_B)
mann_whitney

# Boxplot for Mann–Whitney
ggplot(independent_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Mann–Whitney U test: Group A vs B", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel2")

# -------------------------------
# 4. Wilcoxon signed-rank test (Non-parametric, paired)
# -------------------------------
wilcoxon_signed <- wilcox.test(before, after, paired = TRUE)
wilcoxon_signed

# Paired plot for Wilcoxon
ggplot(paired_df_long, aes(x = Time, y = Value, group = Subject)) +
  geom_line(color = "gray") +
  geom_point(aes(color = Time), size = 3) +
  labs(title = "Wilcoxon Signed-Rank Test: Before vs After", y = "Value") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

### Non-Parametric

#### Mann–Whitney U (NOT DONE)

```{r}
# -----------------------------------------
# ### Non-Parametric Tests
# -----------------------------------------

# Simulate data for independent groups (Mann–Whitney U)
set.seed(123)
group_ind <- rep(c("A","B"), each=20)
y_ind <- c(rnorm(20, 50, 10), rnorm(20, 55, 10))

# Boxplot to visualize
boxplot(y_ind ~ group_ind, col=c("lightblue","lightgreen"),
        main="Independent Groups - Mann–Whitney U Test",
        ylab="Values", xlab="Group")

# Mann–Whitney U test (independent, non-parametric)
mann_whitney_result <- wilcox.test(y_ind ~ group_ind, exact=FALSE)
mann_whitney_result
```

#### Wilcoxon signed-rank

```{r}
# -----------------------------------------
# Simulate data for paired samples (Wilcoxon signed-rank)
set.seed(124)
y_paired_before <- rnorm(15, 50, 5)
y_paired_after  <- y_paired_before + rnorm(15, 3, 4)  # paired differences

# Boxplot to visualize paired differences
boxplot(y_paired_before, y_paired_after, names=c("Before","After"),
        col=c("orange","pink"), main="Paired Samples - Wilcoxon Signed-Rank Test",
        ylab="Values")

# Wilcoxon signed-rank test (paired, non-parametric)
wilcoxon_result <- wilcox.test(y_paired_before, y_paired_after, paired=TRUE)
wilcoxon_result
```

#### quick Environment cleaning

```{r}
rm(list=ls())
```

## Hypothesis testing (Multiple groups):

### parametric One-way ANOVA

```{r}
# -------------------------------
# R Script: Multiple Group Hypothesis Testing
# -------------------------------

# Load libraries
library(ggplot2)
library(dplyr)
library(FSA)       # For Dunn test
library(rcompanion) # For Dunn test adjusted p-values

# -------------------------------
# Example dataset
# -------------------------------
set.seed(123)
group1 <- c(45, 50, 52, 47, 55)
group2 <- c(60, 62, 58, 65, 63)
group3 <- c(55, 53, 50, 57, 59)

# Combine into a dataframe
multi_df <- data.frame(
  Value = c(group1, group2, group3),
  Group = factor(rep(c("Group1", "Group2", "Group3"), each = 5))
)

# -------------------------------
# 1. Visualize the groups
# -------------------------------
ggplot(multi_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Multiple Groups: Distribution of Values", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

# -------------------------------
# 2. Parametric: One-way ANOVA
# -------------------------------
anova_result <- aov(Value ~ Group, data = multi_df)
summary(anova_result)

# -------------------------------
# 2a. Post-hoc: Tukey HSD
# -------------------------------
tukey_result <- TukeyHSD(anova_result)
tukey_result
```

#### non-parametric: Kruskal–Wallis Test

```{r}
# -------------------------------
# 3. Non-parametric: Kruskal–Wallis Test
# -------------------------------
kruskal_result <- kruskal.test(Value ~ Group, data = multi_df)
kruskal_result

# -------------------------------
# 3a. Post-hoc: Dunn Test
# -------------------------------
dunn_result <- dunnTest(Value ~ Group, data = multi_df, method="bonferroni")
dunn_result
```

#### Anova vs Kruskal-Wallis

```{r}
# -------------------------------
# Compare ANOVA vs Kruskal-Wallis Visualizations
# -------------------------------

# Load libraries
library(ggplot2)
library(gridExtra)  # for arranging multiple plots

set.seed(123)

# -------------------------------
# 1. ANOVA dataset (normal)
# -------------------------------
anova_group1 <- rnorm(10, mean = 50, sd = 5)
anova_group2 <- rnorm(10, mean = 55, sd = 5)
anova_group3 <- rnorm(10, mean = 60, sd = 5)

anova_df <- data.frame(
  Value = c(anova_group1, anova_group2, anova_group3),
  Group = factor(rep(c("Group1", "Group2", "Group3"), each = 10))
)

anova_plot <- ggplot(anova_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2, color = "black") +
  labs(title = "One-way ANOVA (Normal Data)", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

# -------------------------------
# 2. Kruskal-Wallis dataset (skewed)
# -------------------------------
kruskal_groupA <- rexp(10, rate = 0.1) + 40
kruskal_groupB <- rexp(10, rate = 0.1) + 50
kruskal_groupC <- rexp(10, rate = 0.1) + 60

kruskal_df <- data.frame(
  Value = c(kruskal_groupA, kruskal_groupB, kruskal_groupC),
  Group = factor(rep(c("GroupA", "GroupB", "GroupC"), each = 10))
)

kruskal_plot <- ggplot(kruskal_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2, color = "black") +
  labs(title = "Kruskal-Wallis (Skewed Data)", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel2")

# -------------------------------
# Arrange both plots vertically
# -------------------------------
grid.arrange(anova_plot, kruskal_plot, ncol = 1)

# -------------------------------
# Steps to Showcase Skewness
# -------------------------------
      # 1. Histogram
      # Shows the shape of the distribution
      # Skewed data will have a long tail to one side

hist(kruskal_df$Value,
     breaks = 10,
     main = "Histogram of Kruskal-Wallis Data",
     xlab = "Value",
     col = "lightblue")

      # 2. Density Plot
      # Smoothed version of histogram
      # Makes skew more obvious

plot(density(kruskal_df$Value),
     main = "Density Plot of Kruskal-Wallis Data",
     xlab = "Value",
     ylab = "Density",
     col = "blue", lwd = 2)
polygon(density(kruskal_df$Value), col = rgb(0.2,0.5,0.8,0.3))

      # 3. Q–Q Plot
      # Compares your data to a theoretical normal distribution
      # For skewed data, points curve away from the diagonal

qqnorm(kruskal_df$Value, main = "Q-Q Plot of Kruskal-Wallis Data")
qqline(kruskal_df$Value, col = "red")

      # 4. Skewness Statistic
      # Quantifies the asymmetry of the distribution
      # Positive skew → right tail longer
      # Negative skew → left tail longer
# Install e1071 if not already installed
# install.packages("e1071")
library(e1071)

# skewness_value <- skewness(kruskal_df$Value)
cat("Skewness =", round(skewness(kruskal_df$Value), 3), "\n")
cat("Skewness =", round(skewness(anova_df$Value), 3), "\n")
      # Interpretation:
      # skewness > 0 → right-skewed (common for Kruskal-Wallis example)
      # skewness ≈ 0 → approximately symmetric (normal)
      # skewness < 0 → left-skewed

      # 5. Optional: Violin Plot
library(ggplot2)

# Violin plot with individual points
ggplot(kruskal_df, aes(x = Group, y = Value, fill = Group)) +
  geom_violin(trim = FALSE, alpha = 0.6) +   # Show full density, not trimmed
  geom_jitter(width = 0.1, size = 2, color = "black") +  # Show individual points
  stat_summary(fun = median, geom = "point", shape = 23, size = 3, fill = "white") +  # median
  labs(title = "Violin Plot: Kruskal-Wallis Groups", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel2")

```

### One-way ANOVA Visualization

Dataset: Normally distributed values for 3 independent groups

Why this works for ANOVA:

Boxplots clearly show the mean and spread of each group.

Jittered points show the individual observations.

Parametric ANOVA assumes normally distributed data, so this visualization highlights the mean differences and variance.

```{r}
# Load libraries
library(ggplot2)

set.seed(123)
# Normally distributed data for parametric ANOVA
group1 <- rnorm(10, mean = 50, sd = 5)
group2 <- rnorm(10, mean = 55, sd = 5)
group3 <- rnorm(10, mean = 60, sd = 5)

anova_df <- data.frame(
  Value = c(group1, group2, group3),
  Group = factor(rep(c("Group1", "Group2", "Group3"), each = 10))
)

# Visualization: Boxplot + jitter for ANOVA
ggplot(anova_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2, color = "black") +
  labs(title = "One-way ANOVA: Comparison of Group Means", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

### Kruskal–Wallis Visualization

Dataset: Skewed/non-normal values for 3 independent groups

Boxplots show median, spread, and skew — important since Kruskal–Wallis compares ranks, not means.

Jittered points highlight outliers and distribution shape, which can affect non-parametric tests.

This type of plot makes it clear why non-parametric tests are preferred for skewed or non-normal data.

```{r}
set.seed(123)
# Skewed data for non-parametric test
groupA <- rexp(10, rate = 0.1) + 40
groupB <- rexp(10, rate = 0.1) + 50
groupC <- rexp(10, rate = 0.1) + 60

kruskal_df <- data.frame(
  Value = c(groupA, groupB, groupC),
  Group = factor(rep(c("GroupA", "GroupB", "GroupC"), each = 10))
)

# Visualization: Boxplot + jitter for Kruskal-Wallis
ggplot(kruskal_df, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2, color = "black") +
  labs(title = "Kruskal–Wallis Test: Comparison of Group Ranks", y = "Value") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel2")
```

#### quick Environment cleaning

```{r}
rm(list=ls())
```

## Effect Size

### Fold-change

```{r}
group1 <- c(10,12,9,11,10)
group2 <- c(15,16,14,18,17)

fc <- mean(group2) / mean(group1)
log2fc <- log2(fc)

fc
log2fc
```

### Cohen’s d

```{r}
x <- rnorm(20, mean=10, sd=2)
y <- rnorm(20, mean=12, sd=2)

d <- (mean(x) - mean(y)) /
     sqrt((var(x) + var(y))/2)

d
# or write it in as a function below
cohen_d <- function(x, y) {
  nx <- length(x)
  ny <- length(y)
  pooled_sd <- sqrt(((nx-1)*var(x) + (ny-1)*var(y))/(nx+ny-2))
  (mean(x) - mean(y))/pooled_sd
}

cohen_d(x, y)

```

### Hedges’ g

```{r}
n <- length(x) + length(y)

g <- d * (1 - (3 / (4*n - 9)))

g
```

### Bonferroni correction

```{r}
pvals <- runif(10)
cat("lets first check the raw p-values")
pvals
cat("now see how the bonferroni correction adjusts them")
p.adjust(pvals, method="bonferroni")


```

### Benjamini–Hochberg FDR

```{r}
cat("now see how the Benjamini–Hochberg FDR correction adjusts them")
p.adjust(pvals, method="BH")
```

### q-values

```{r}
cat("now see how the q-value correction adjusts them")
p.adjust(pvals, method="fdr")
```

## Confidence Intervals

```{r}
t.test(x, y)

t.test(x, y)$conf.int
```
