---
title: 'Week 5: Data Visualization'
author: "Reika Masuda"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Graphic Systems in R – Examples

## 1. Base Graphics

*Simple, fast “pen on paper” plotting*

```{r}
# Scatterplot
x <- rnorm(50)
y <- 2*x + rnorm(50)

plot(x, y, main="Base R Scatterplot", xlab="X", ylab="Y", pch=16, col="blue")

# Add regression line (layered, cannot modify previous)
abline(lm(y ~ x), col="red", lwd=2)
```

*compared to ggplot2*

```{r}
library(ggplot2)
df <- data.frame(x, y)
ggplot(df, aes(x=x, y=y)) +
  geom_point(color="blue", size=2) +
  geom_smooth(method="lm", color="red", se=FALSE) +
  labs(title="ggplot2 Scatterplot", x="X", y="Y") +
  theme_minimal()

```

## 2. Grid Graphics

*Low-level control using graphical objects (grobs)*

```{r}
library(grid)

# Create a rectangle and text grob
grid.newpage()
grid.rect(width = 0.5, height = 0.5, gp = gpar(fill = "lightblue"))

grid.text("Grid Graphics",
          x = 0.5, y = 0.5,
          gp = gpar(fontsize = 16))
```

## 3. Lattice Graphics

*Conditioned (multi-panel) plots*

```{r}
library(lattice)

# Use built-in dataset
xyplot(mpg ~ wt | factor(cyl),
       data = mtcars,
       main = "Lattice: MPG vs Weight by Cylinders",
       col = "darkgreen",
       pch = 19)

```

## 4. ggplot2

*Layered grammar of graphics*

```{r}
# library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ cyl) +
  labs(title = "ggplot2: MPG vs Weight by Cylinders")
```

## 5. Interactive Graphics (plotly)

*Interactive version of ggplot*

```{r}
library(plotly)

p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

ggplotly(p)
```

### add more features in the interactives

```{r}
p1 <- mtcars%>%
  tibble::rownames_to_column("model") %>%
  ggplot(aes(x = wt, y = mpg,ID = model)) +
  geom_point()

ggplotly(p1)
```

*clean up environment*

```{r}
rm(list = ls())
```

### tidyr pivot_longer(), facet_wrap()

```{r}
# library(ggplot2)
library(tidyr)
library(dplyr)

# Select numeric measurement columns
df <- iris %>% select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, Species)

# Pivot longer to long format
df_long <- df %>%
  pivot_longer(
    cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
    names_to = "Measurement",
    values_to = "Value"
  )

# Plot with facets
ggplot(df_long, aes(x = Species, y = Value)) +
  geom_point(aes(color = Species), size = 2) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "Faceted iris Measurements", x = "Species", y = "Value") +
  theme_minimal()
```

### compare multiple plotting geom\_\*

```{r}
ggplot(df_long, aes(x = Species, y = Value)) +
  geom_point(aes(color = Species), size = 2) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "geom_point", x = "Species", y = "Value") +
  theme_minimal()->p1 
plot(p1)

ggplot(df_long, aes(x = Species, y = Value)) +
  geom_jitter(aes(color = Species),width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "geom_jitter", x = "Species", y = "Value") +
  theme_minimal()->p2 
plot(p2)


ggplot(df_long, aes(x = Species, y = Value)) +
  geom_boxplot(aes(color = Species)) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "geom_boxplot", x = "Species", y = "Value") +
  theme_minimal()->p3 
plot(p3)

ggplot(df_long, aes(x = Species, y = Value)) +
  geom_boxplot(aes(color = Species)) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "geom_boxplot", x = "Species", y = "Value") +
  theme_minimal()->p4 
plot(p4)

ggplot(df_long, aes(x = Species, y = Value)) +
  geom_violin(aes(color = Species)) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "geom_violin", x = "Species", y = "Value") +
  theme_minimal()->p5 
plot(p5)
```

### Combine half-violin, boxplot, and jitter in one plot

```{r}
library(gghalves)
ggplot(df_long, aes(x = Species, y = Value, fill = Species)) +
  
  # LEFT
  geom_half_violin(side = "l", alpha = 0.6, color = NA) +
  geom_half_boxplot(side = "l", width = 0.3, outlier.shape = NA, alpha = 0.8) +
  geom_jitter( size = 1, alpha = 0.5,position = position_nudge(x = +0.1))+
  facet_wrap(~Measurement, scales = "free_y") +
  
  labs(title = "Half-violin + boxplot + points (left-right split)",
       x = "Species", y = "Value") +
  
  theme_minimal() +
  theme(legend.position = "none")
  

```

### Arrange multiple grobs on a page using gridExtra: grid.arrange()

```{r}
library(gridExtra)
library(ggpubr)

# check what the grid.arrange() function does and how to use it to arrange multiple ggplot objects on a single page.
?grid.arrange

# Previously created ggplot objects: p1, p2, p3, p4, p5 you can arrange them in a 2-row, 3-column layout (with one empty space) using grid.arrange() as follows:
grid.arrange(p1, p2, p3, p4, p5, nrow = 2, ncol = 3)

# I dont want all the lengends to be shown, so I will remove the legends from p1 to p5 and put the legend as one ggplot object in the last position of the grid.arrange() layout. To do this, you can use the theme() function to remove the legends from each plot and then create a separate plot for the legend. Here's how you can do it:

p_legend <- get_legend(p1)
# p_legend <- ggpubr::get_legend(p1) ## if you havent called ggpubr yet, you may need to specify the package for get_legend()

grid.arrange(p1+theme(legend.position = "none"), 
             p2+theme(legend.position = "none"),
             p3+theme(legend.position = "none"),
             p4+theme(legend.position = "none"),
             p5+theme(legend.position = "none"),
             p_legend,nrow = 2, ncol = 3)


# is the figure too small for the display?
# change the top of this chunck to ... {r,fig.height=10,fig.width=15} and run it again
```

# Visualizing Statistical Comparisons: Iris Dataset Example

```{r}
# library(ggplot2)
# library(ggpubr)

# check what the stat_compare_means() function does and how to use it to add statistical comparisons to ggplot2 plots.
?stat_compare_means

# Basic boxplot + jitter
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +

  # Global test
  stat_compare_means(method = "anova", label.y = 9.5) +

  # Pairwise comparisons
  stat_compare_means(
    aes(label = ..p.signif..),
    method = "t.test",
    comparisons = list(
      c("setosa", "versicolor"),
      c("setosa", "virginica"),
      c("versicolor", "virginica")
    ),
    label.y = c(8.5, 9, 9.2)
  ) +

  labs(
    title = "Sepal.Length by Species with ANOVA & Pairwise Comparisons",
    x = "Species",
    y = "Sepal.Length"
  ) +
  theme_minimal()
```

## 1. t-test (Two groups)

Use case: Compare the means of two independent groups Example: Compare Sepal.Length between setosa vs versicolor

```{r}
# library(ggpubr)

ggplot(subset(iris, Species %in% c("setosa","versicolor")),
       aes(x = Species, y = Sepal.Length, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_compare_means(method = "t.test") +
  theme_minimal()
```

## 2. ANOVA (More than two groups)

Use case: Compare means across three or more independent groups Example: Compare Sepal.Length across all species (setosa, versicolor, virginica)

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_compare_means(method = "anova") +
  theme_minimal()
```

## 3. Kruskal-Wallis (Non-parametric ANOVA)

Use case: Compare distributions across multiple groups when normality is not assumed Example: Sepal.Width across all species

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_compare_means(method = "kruskal.test") +
  theme_minimal()
```

## 4. Wilcoxon / Mann-Whitney U Test (Non-parametric t-test)

Use case: Compare two independent groups without assuming normality Example: Compare Petal.Length between setosa and versicolor

```{r}
ggplot(subset(iris, Species %in% c("setosa","versicolor")),
       aes(x = Species, y = Petal.Length, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_compare_means(method = "wilcox.test") +
  theme_minimal()
```

### Faceted plot with statistical comparison

```{r,fig.width=8,fig.height=8}
# library(ggplot2)
# library(ggpubr)
# library(tidyr)

# Convert iris to long format
df_long <- iris %>%
  pivot_longer(cols = starts_with("Sepal") | starts_with("Petal"),
               names_to = "Measurement",
               values_to = "Value")

ggplot(df_long, aes(x = Species, y = Value, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  stat_compare_means(method = "anova", label.y = max(df_long$Value) + 0.2) +
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "Faceted Statistical Comparisons Across Iris Measurements",
       x = "Species", y = "Value") +
  theme_minimal()
```

### Faceted Statistical Comparison with Pairwise Tests

ns: p \> 0.05

\*: p \<= 0.05

\*\*: p \<= 0.01

\*\*\*: p \<= 0.001

\*\*\*\*: p \<= 0.0001

```{r,fig.width=8,fig.height=8}
comparisons <- list(
  c("setosa", "versicolor"),
  c("setosa", "virginica"),
  c("versicolor", "virginica")
)

ggplot(df_long, aes(x = Species, y = Value, color = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  
  # pairwise comparisons using compare_means()
  stat_compare_means(
    method = "t.test",
    comparisons = comparisons,
    label = "p.signif"
  ) +
  
  facet_wrap(~Measurement, scales = "free_y") +
  labs(title = "Faceted Pairwise Comparisons Across Iris Measurements",
       x = "Species", y = "Value") +
  theme_minimal()+
  theme(legend.position = "bottom")
```
