---
title: "Week 2: Workshop"
author: "Megan Collins"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# library(knitr)
# library(dplyr)
# library(readxl)

# Set your working directory
 setwd("C:/Users/colli/git/BIO513/Week 2")
```

# Data file Format

```{r}
data<-mtcars
write.csv(data,"data.txt")
write.csv(data,"data.csv")
write.csv2(data,"data2.csv")
write.table(data,"data.tsv")
rm(data)
```

*Lets open your folder and duplicate data.txt (which is comma separated) replace all the "," with "\|" save as data2.txt*

## Plain Text .txt style

```{r,Plain Text .txt style}
# Plain text has No specific structure nor delimiter

# Let's read in data.txt with read.csv()
read.csv("data.txt")
# Now lets read in using delim()
read.delim("data.txt", sep = ",")

# What about data2.txt ?
# Can you read in using read.csv()?
read.csv("data2.txt")
# That looks a bit strange isn't it?
# Lets specify the delimer "|" still using read.csv()
read.csv("data2.txt", sep = "|")
# How about read in with read.delim()?
read.delim("data2.txt", sep = "|")

```

## Comma-separated values .csv style

```{r, comma-separated values .csv style}

# Lets read in data.csv using read.csv()
read.csv("data.csv", header = TRUE, sep = ",")

# What will happen if we say header = FALSE?
read.csv("data.csv", header = FALSE, sep = ",")

```

## Semicolon-Separated values .csv2 style

```{r,Semicolon-Separated values .csv2 style,error=TRUE,warning=FALSE}

# can we read in data2.csv (csv2 style) with read.csv()?
read.csv("data2.csv")
# Did you get the following error?
# Error in read.table(file = file, header = header, sep = sep, quote = quote,  :
#   more columns than column names

#How about now?
read.csv("data2.csv", sep = ";")

# Or using proper function read.csv2()?
read.csv2("data2.csv")

```

## Tab separated values .tsv style

```{r,tab separated values .tsv style}
read.table("data.tsv")
```

### Bonus : excel sheets

You will need "readxl" package

```{r,eval=FALSE}
# install.packages("readxl")    # remember? you only need to do this once

library(readxl)
# You have a .xlx file called "datasets.xlsx" In this file consists of 5 sheets: "mtcars", "iris", "Titanic", "airquality", and "ChickWeight". They are all datasets that are available in base R

# if you don't define which sheet, read_excel() read in the first sheet
read_excel("datasets.xlsx")

# Lets read in "Titanic"
read_excel("datasets.xlsx", sheet = "Titanic")

```

# Explore datasets

## Iris dataset

```{r}

# Read in iris sheets of the datasets.xlsx and name this dataset as "iris"
#iris <- read_excel("datasets.xlsx", sheet = "iris")
iris <- iris
cat("What data type is iris?")
class(iris)

cat("What are the dimension?")
dim(iris)

cat("What are the variables?")
colnames(iris)

# Let's have a peak of iris data
head(iris)
cat("There are 5 columns in iris data.frame.",
    "\n",
    "All apart from Species are numeric")

cat("What is the range of Sepal Length?")
range(iris$Sepal.Length)

cat("What are the minimum, mean, median, and maximum of Sepal Width?")
min(iris$Sepal.Width)
mean(iris$Sepal.Width)
median(iris$Sepal.Width)
max(iris$Sepal.Width)
## Or use summary()
summary(iris$Sepal.Width)

cat("Can we visualize Petal Length using histogram?")
hist(iris$Petal.Length)

cat("How many different species in this iris dataset?")
length(unique(iris$Species))

cat("What are they?")
unique(iris$Species)

cat("How many of each species do we have?")
table(iris$Species)


```

# Which function

```{r}
cat("Which row index are representing versicolor in iris dataset?")

which(iris$Species == "versicolor")

# Lets check
iris[which(iris$Species == "versicolor"), ]

# summary(iris$Petal.Width[which(iris$Species=="versicolor")])
cat(
  "Which row index are representing versicolor",
  "\n",
  "and also its Petal width greater than 1.3 in iris dataset?"
)

which(iris$Species == "versicolor" & iris$Petal.Width > 1.3)

# Lets check
iris[which(iris$Species == "versicolor" & iris$Petal.Width > 1.3), ]


```

## using %in% for which()

*which(A %in% B)*

```{r}
# Which iris row index are corresponding to "versicolor" and  "virginica" Species?
which(iris$Species %in% c("versicolor", "virginica"))
which(iris$Species=="versicolor" | iris$Species=="virginica")
# Or in other words, which are NOT "setosa" ?
which(iris$Species != "setosa")
```

# Match function

*match(reference , data to flip)*

```{r}
# Lets make a two dummy string
a<-c("coffee","matcha","juice","milkshake","wine")
b<-c("tea","coffee","water","wine","juice")

# If we want to match the order of a in b: meaning the order of a stays and change the order of b
match(a,b)
# lets check 
b[match(a,b)]

# can we do the oppsite?
match(b,a)
a[match(b,a)]

```

## Challenge #1: Can you find the common entires within a and b, and match the order of both subset?

```{r}
# Which entries in a is also found in b?
idx<-which(a %in% b)
# let's name the subset a1 
a1<-a[idx]
rm(idx)
# Which entries in b is also found in a?
idx<-which(b %in% a)

# let's name the subset b1 
b1<-b[idx]

# do a1 and b1 has the same length?
length(a1)==length(b1)

# lets re-order the b1 to match with a1
## checking a1
a1
## checking b1
b1
## checking if the order of a1 and a2 are really different to begin with
a1==b1
table(a1==b1)
## inorder to match the order of b1 as a1, the array must be reorder as:
match(a1,b1)
## lets check
b1[match(a1,b1)]
## Are they really the same order?
a1==b1[match(a1,b1)]
## we can overwrite b1 with matched order
b1<-b1[match(a1,b1)]

```

## Challenge #2: Can you find the common entires match two data.frame using ID column and make one data.frame "data3"?

```{r}
data1 <- data.frame(ID = paste0("ID_", seq(1:10)), randu[1:10, ]) # randu is another random dataset with 3 column x,y, and z

data2 <- data.frame(
  ID = paste0("ID_", sample(1:15)),
  Age = round(runif(15, min = 20, max = 50), 0),
  Sex = c(rep("F", 8), rep("M", 7))
)
```

### Solution to Challenge #2:

```{r}
data1 <- data.frame(ID = paste0("ID_", seq(1:10)), 
                    randu[1:10, ]) # randu is another random dataset with 3 column x,y, and z

data2 <- data.frame(
  ID = paste0("ID_", sample(1:15)),
  Age = round(runif(15, min = 20, max = 50), 0),
  Sex = c(rep("F", 8), rep("M", 7))
)

              ### Solution ###

# which data1 ID are in data2 ID?
which(data1$ID %in% data2$ID)
data1 <- data1[which(data1$ID %in% data2$ID), ]
# which data2 ID are in data1 ID?
which(data2$ID %in% data1$ID)
data2 <- data2[which(data2$ID %in% data1$ID), ]
# Now you should see both data1 and data2 have 10 observation in total
# Lets keep data1 ID order as a reference (do not move) and change the order of data2 dataset

# match the order of data2 as data1
match(data1$ID,data2$ID)
data2<-data2[match(data1$ID,data2$ID),]

# combine data1 and data2 using cbind() 
data3<-cbind(data1,data2[,-1])

data3
```

### Cleaning up the Environment *rm(list = ls())*

```{r}
# Environment is getting a bit messy.. lets clean up 
rm(list = ls())
```

# dplyr

```{r}
# install.packages("dplyr") # you only need to do this once
library(dplyr)
# Lets use mtcars data for now
```

## select()

```{r}
colnames(mtcars)

# let's select a column "mpg", "hp", "gear"

mtcars %>% select(mpg, hp, gear)

# let's select a column start with "d" on the column name: "disp", "drat"
mtcars %>% select(starts_with("d"))

# let's select a column ends with "t" on the column name:  "drat", "wt"
mtcars %>% select(ends_with("t"))

```

## filter()

```{r}
# let's subset (*filter*) mtcars data where mpg is greater than 15
mtcars %>% filter(mpg > 15)

# Try filtering mpg >= 13.0 and hp > 100
mtcars %>% filter(mpg >= 13 & hp < 100)

# How about disp < 140 OR vs = 1?
mtcars %>% filter(disp < 140 | vs == 1)

```

## Mutating Joins

We can solve the challenge #2 using dplyr functions

Lets explore what each joining function can do

```{r}
data1 <- data.frame(ID = paste0("ID_", seq(1:10)), randu[1:10, ]) # randu is another random dataset with 3 column x,y, and z

data2 <- data.frame(
  ID = paste0("ID_", sample(1:15)),
  Age = round(runif(15, min = 20, max = 50), 0),
  Sex = c(rep("F", 8), rep("M", 7))
)
```

### Inner Join: inner_join()

An inner_join(x,y) only keeps observations from x that have a matching key in y.

```{r}
inner_join(data1, data2, by = "ID")
```

### Outer Join: left_join()

keeps all observations in x.

```{r}
left_join(data1, data2, by = "ID")

# What if we swap the order?
left_join(data2, data1, by = "ID")

```

### Outer Join: right_join()

keeps all observations in y.

```{r}
right_join(data1, data2, by = "ID")

# What if we swap the order?
right_join(data2, data1, by = "ID")
```

### Outer join: full_join()

keeps all observations in x and y.

```{r}
full_join(data1, data2)

full_join(data2, data1)
# Can you spot the difference?
```

## Base R: merge()

```{r}

merge(data1, data2)

merge(data2, data1)
# Can you spot the difference?
```

# Date and Time handling

```{r}
# Example random dates for R exercise
dates <- c(
  "2026-03-01",       # ISO format
  "03/01/2026",       # US format
  "01-03-2026",       # European format
  "Mar 1, 2026",      # Month name
  "March 01 2026",    # Full month
  "2026/03/01",       # Slash ISO
  "01-Mar-26",        # Short year
  "1 Mar 2026",       # Day Month Year
  "2026-3-1",         # Missing leading zeros
  "03-1-2026",        # Mixed formatting
  "2026-03-32",       # Typo: invalid day
  "Feb 30, 2026",     # Typo: non-existent day
  "2026/13/01",       # Typo: invalid month
  "Mar-01-202",       # Typo: year missing digit
  "2026-03-0a"        # Typo: invalid character
)

# Convert valid dates to Date class (will produce NA for invalid formats)
as.Date(dates, tryFormats = c("%Y-%m-%d", "%m/%d/%Y", "%d-%m-%Y", "%b %d, %Y", "%d %b %Y"))
```

## making dummy Date.xlsx (and save as .xlsx file)

```{r}

################### mixed_dates (Collection Date ) ################### 
set.seed(123) # for reproducibility

# Generate 30 random valid dates
valid_dates <- as.Date("2025-01-01") + sample(0:730, 30, replace = TRUE) # random dates over 2 years

# Create mixed formats
random_dates <- sapply(valid_dates, function(d) {
  fmt <- sample(1:5, 1)
  if(fmt == 1) format(d, "%Y-%m-%d")       # ISO
  else if(fmt == 2) format(d, "%m/%d/%Y")  # US
  else if(fmt == 3) format(d, "%d-%m-%Y")  # EU
  else if(fmt == 4) format(d, "%b %d, %Y") # Month name short
  else format(d, "%d %b %Y")               # Day Month Year
})

# Add 10 Excel-style serial numbers (dates as numbers)
# !!! NOTE: simply converting a Date to numeric in R gives the number of days since 1970-01-01, whereas Excel stores dates as days since 1899-12-30 (for Windows Excel) 
excel_dates <- as.character(as.numeric(valid_dates[1:10] - as.Date("1899-12-30")))

# Add 10 typo / messy dates
typo_dates <- c(
  "2026-02-30", "13/01/2026", "Feb 31, 2026", "03-15-26", 
  "2025/2/29", "Mar-01-2a6", "2026-0a-05", "1/1/26", 
  "Jul 32 2025", "2025/13/01"
)

# Combine all
mixed_dates <- c(random_dates, excel_dates, typo_dates)

# Shuffle
set.seed(456)
mixed_dates <- sample(mixed_dates)

# Display
as.data.frame(mixed_dates)
################### formatted_dates (DOB) ################### 
set.seed(789)  # for reproducibility

# Generate 50 random dates from 1970-01-01 to 1999-12-31
start_date <- as.Date("1970-01-01")
end_date <- as.Date("1999-12-31")
random_dates <- start_date + sample(0:(as.integer(end_date - start_date)), 50, replace = TRUE)

# Convert to different common formats
formatted_dates <- sapply(random_dates, function(d) {
  fmt <- sample(1:4, 1)
  if(fmt == 1) format(d, "%Y-%m-%d")       # ISO
  else if(fmt == 2) format(d, "%m/%d/%Y")  # US
  else if(fmt == 3) format(d, "%d-%m-%Y")  # EU
  else format(d, "%b %d, %Y")              # Month name short
})

# Shuffle just in case
set.seed(321)
formatted_dates <- sample(formatted_dates)

# Show the dates
as.data.frame(formatted_dates)
################### Construct Dummy Metadata ################### 

Date = data.frame(
  ID = seq(1,50,1),
  DOB = formatted_dates,
  CollectionDate = mixed_dates,
  Sex = sample(c("M", "F"), 50, replace = TRUE),
  height = round(runif(50, min = 150, max = 200), 1),
  weight = round(runif(50, min = 50, max = 100), 1),
  Group = sample(c("Control", "Treatment"), 50, replace = TRUE),
  hsCRP = round(runif(50, min = 0.1, max = 10), 2),
  HbA1c = round(runif(50, min = 4.0, max = 10.0), 1),
  Trigricerides = round(runif(50, min = 50, max = 300), 1),
  Cholesterol = round(runif(50, min = 100, max = 250), 1),
  stringsAsFactors = FALSE
)

################### Add few mistake s################### 
# introduce some typo in Sex (some in lower case, some written as male, female)
Date$Sex[c(11,13)]<-"female"
Date$Sex[c(22,33)]<-"male"
Date$Sex[c(24,35)]<-c("f","m")
table(Date$Sex)
     # f      F female      m      M   male 
     # 1     20      2      1     24      2 

# lets make one of the height in m instead of cm
Date$height[5]<-Date$height[5]/100

# Lets swap height and weight for one of the observation
temp<-Date$height[10]
Date$height[10]<-Date$weight[10]
Date$weight[10]<-temp
rm(temp)

# introduce some typo in Group (some in lower case, some written as con, treat)
Date$Group[c(12,14)]<-"con"
Date$Group[c(25,36)]<-"treat"

# lets add NA in few of the hsCRP and HbA1c
Date$hsCRP[c(15,26)]<-NA
Date$HbA1c[c(16,27)]<-NA
# save it as .xlsx
writexl::write_xlsx(Date, "Date.xlsx")
```

### Cleaning up the Environment *rm(list = ls())*

```{r}
# Environment is getting a bit messy.. lets clean up 
rm(list = ls())
```

## read in the Date.xlsx file using read_excel() function

```{r}
library(readxl)
data<-read_excel("Date.xlsx")
# Let's have a look
data
```

## Excel-style serial numbers using as.Date() function

```{r}
as.Date(45956, origin = "1899-12-30")
```

# Lubridate

lubridate is basically the superhero of handling dates and times in R. Before it, working with dates could be a real headache—parsing strings, extracting months, adding days… it was a lot of manual work. lubridate makes all of that super smooth.

```{r}
# install.packages("lubridate") # you only need to do this once
library(lubridate)

      ### Example of parsing dates with lubridate ###
# Different formats
ymd("20260301")          # "2026-03-01"
mdy("03/01/2026")        # "2026-03-01"
dmy("01-03-2026")        # "2026-03-01"

# With times
ymd_hms("2026-03-01 14:30:00")  # "2026-03-01 14:30:00 UTC"
mdy_hm("03/01/2026 14:30")      # "2026-03-01 14:30:00 UTC" 
```

## Cleaning DOB

```{r}
DOB<-data$DOB
# Parse DOB using lubridate parse_date_time()

DOB_parsed<-parse_date_time(DOB, orders = c("ymd", "mdy", "dmy", "b d, Y"))

data.frame(original = DOB, parsed = DOB_parsed)
```
